### 2015-07-06 narrative/data-reduction.tex

In this section we describe some common methods to reduce and re-organize data.
There are many possibilities and the best choice will generally vary from case
to case but, once settled, will not change from run to run. It will be
important understand when the time and resources to required to identify and
perform the ``best methods'' outweighs the eventual gains. Another critical
research question concerns the quantification and control of the loss of
information resulting from refactoring the data and using a reduced 
dataset.

@@

Application knowledge means one can sometimes achieve drastically superior
reduction in the total data compared with what one might achieve otherwise.
However, in the absence of such knowledge, we must equip the SSIO layers with
generic data reduction techniques and re-organization techniques.
Nevertheless, certain basic information on the semantics of the data is needed 
and must be supported by the overall infrastructure.

Another important consideration is knowledge regarding typical access patterns
for data in a given application. Scientific applications often read and analyze
data in a structured fashion and failure to respect this pattern, or to exploit
it through prefetching, can significantly increase the overall execution time
of the application. Accurate prefetching can only be achieved with access
pattern analysis. This requires an SSIO layer which can prefetch data in a
asynchronous fashion to take the bytes on the file.

There are many methods to re-organize the data, and to reduce the data and in
this section we will list many different known methods.  We then list our
different approaches to address the many research challenges using many well
known methods, and a new method that we have been working on, known as an
``auditor'' which can be used ``in situ'' with the calculation during the data
generation, and then used to dramatically reduce the impact on the SSIO layer
for both reading and writing.

\paragraph{State of the art:} {\bf Data Reduction Techniques} Data compression
broadly consists of three basic steps: (1) Pre-processing to extract as much
structure as feasible. This step is variously referred to as filtering,
preconditioning, and pre- processing but in all cases, the objective is to
represent the data in a form that becomes more amenable to compression.  (2)
Removal of redundancy such as omission of duplicates etc.  Finally, (3) the
resulting data is compressed using entropy coding. The level of compression
that can be achieved in the final step is limited by the Shannon information
content of the data \cite{Gray:book}, and many standard tools are available for
efficient and perfect entropy coding. The potential for compression through
removal of redundancy is generally minor. The main scope for achieving
significant compression (a compression rate of two or better say) therefore
lies in the pre-processing step.

Research in the construction of pre-processing techniques is still in its
infancy, especially in context of exascale simulations. Techniques that have
been advocated range from simple difference filtering, bit-sorting, prefix
transformation through to simply reordering the data.  What these, and other
catchall methods, fail to exploit is a priori information on the nature and
provenance of the data.

As we have already observed, the incorporation of a priori information of this
type is vital if one is to obtain significant data reduction. The use of an
Auditor as a pre-processor exploits the wide (and increasing) cost gap between
CPU operations versus storage-based operations. The preprocessor would ideally
be employed in situ and is used in a similar fashion to pre-processors in
standard compression approaches such as FPC \cite{BurtscherFPC} in which an XOR
operation against the previous data is used as a pre-processor.

>>>>

In this section we describe some common methods to reduce and re-organize data.
There are many possibilities and the best choice will generally vary from case
to case but, once settled, will not change from run to run. It will be
important to understand when the time and resources required to identify and
perform the ``best methods'' outweighs any eventual gains. Another critical
research question concerns the quantification and control of the loss of
information resulting from refactoring the data and using a reduced 
dataset.

@@

Application knowledge means one can sometimes achieve drastically superior
reduction in the total data compared with what one might achieve otherwise.
However, in the absence of such knowledge, we must equip the SSIO layers with
generic data reduction techniques and re-organization techniques.
Nevertheless, certain basic information on the semantics of the data is needed
and must be supported by the overall infrastructure.  Likewise, another
important consideration is knowledge regarding typical access patterns for data
in a given application. Scientific applications often read and analyze data in
a structured fashion and failure to respect this pattern, or to exploit it
through prefetching, can significantly increase the overall execution time of
the application. Accurate prefetching can only be achieved with access pattern
analysis. This requires an SSIO layer which is able to prefetch data in an
asynchronous fashion to take the bytes on the file.

\paragraph{State of the art:} {\bf Data Reduction Techniques} Data compression
broadly consists of three basic steps: (1) Pre-processing to extract as much
structure as feasible. This step is variously referred to as filtering,
preconditioning, and preprocessing but in all cases, the objective is to
represent the data in a form that becomes more amenable to compression.  (2)
Removal of redundancy such as omission of duplicates etc.  Finally, (3) the
resulting data is compressed using entropy coding. The level of compression
that can be achieved in the final step is limited by the Shannon information
content of the data \cite{Gray:book}, and many standard tools are available for
efficient and perfect entropy coding. The potential for compression through
removal of redundancy is generally minor. The main scope for achieving
significant compression (a compression rate of two or better say) therefore
lies in the pre-processing step.

Research in the construction of pre-processing techniques is still in its
infancy, especially in context of exascale simulations. Techniques that have
been advocated range from simple difference filtering, bit-sorting, prefix
transformation through to simply reordering the data.  What these, and other
catch-all methods, fail to exploit is a priori information on the nature and
provenance of the data. Moreover, it is now widely recognized that any
opportunities for {\em in situ} analysis should be exploited as far as
possible. What is less well understood, is how one should incorporate a priori
information and how one can identify possibilities for {\em in situ} analysis.

We shall exploit a priori information in the preprocessing stage through the
use of relatively computationally intensive preprocessors that attempt to
extract structure in the data {\em in situ}. This approach takes advantage of
the increasingly wide cost gap between CPU operations versus storage-based
operations. The preprocessor would generally be employed in situ and is used in
a similar fashion to pre-processors in standard compression approaches such as
FPC \cite{BurtscherFPC} in which an XOR operation against the previous data is
used as a pre-processor.
